---
title: "Assignment 3 - Causal inference"
author: "RF"
date: "2/4/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 3 - Exploring causal inference issues

In this assignment we explore some issues related to multiple regressions (regressions with more than one predictor), and inferred (causal) relations between variables. N.B. the data is simulated (to make sure I know the actual mechanism generating it), but it's based on a real study. So bear with a longish introduction to get into the details of what we are doing and why it is important.
```{r}
library(pacman)
p_load(tidyverse, brms, rethinking, tidybayes)

d <- read.csv("Ass3.csv")

#Releveling
d$Diagnosis <- plyr::revalue(as.character(d$Diagnosis), 
                             c("0"="Controls", "1"="Schizophrenia"))

#ID and Diagnosis as factors
d <- d %>% 
  mutate(
    ID = as.factor(ID),
    Diagnosis = as.factor(Diagnosis)
  )
```


### Altercentric intrusion in schizophrenia

People with schizophrenia often report altered control and distinction of self-other representations: intrusive thoughts, hearing of voices, delusions of mind reading, paranoia, etc (a substantial portion of the psychotic symptoms experienced in schizophrenia). These have been variously attributed to hypermentalizing (over attribution of mental states to others), social impairment (over preoccupation with own thought processes), hyper socialization (inability to inhibit information from others), etc.

The current study investigates 1) whether schizophrenia is indeed related to altered control and distinction of self-other representations, in particular altercentric intrusions (inability to inhibit social information), and 2) whether these are related to the relevant psychotic symptoms. N.B. the actual study also investigates egocentric intrusion, do check the papers below if interested.

The task is a slightly modified version of this: https://www.ncbi.nlm.nih.gov/pubmed/20731512 You look at a picture with some dots visible to you, as well as with a different person with a different set of dots visible to them. The number of dots you see and that the other sees can be the same (congruent condition) or not (incongruent condition). You are tasked to indicate whether a given number (e.g. 3) matches the number of dots you see (and the dots visible to the other person are irrelevant to the task).


The tasks investigates altercentric intrusion: will your reaction time change according to whether the other person is seeing the same amount of dots as you, or not? The idea is that if you correctly inhibit social information, your reaction time should not change, as the information about the other person is not relevant. On the contrary, if you nevertheless use task irrelevant social information, you'll be slower at indicating whether 3 is the right number of dots when the other person sees a different amount of dots than you (conflicting information).
The bigger the difference between RTs in the congruent and incongruent condition the bigger the altercentric intrusion effect.

For each participant you have 6 variables: 1) ID, 2) AltercentricIntrusion (continuous score), 3) Diagnosis (schizophrenia vs. control), 4) VoiceHearing (severity of voice hearing symptoms, continuous score of the severity of the symptom as measured by a clinician), 5) MindReading (severity of delusions of mind reading, continuous score of the severity of the symptom as measured by a clinician); 6) Apathy (severity of lack of motivation in taking care of oneself, from washing to showing up at work, continuous score of the severity of the symptom as measured by a clinician).

The research questions you have to answer are the following:

## First part

Q1.1) Does schizophrenia involved altercentric intrusion? Define model and priors. Test the implications of your priors (prior predictive checks) and if needed adjust them. Run the model. Test the quality of the fitted model (posterior predictive checks). Assess the evidence in favor of an increased altercentric intrusion in schizophrenia. Report the model and the results, including plots.
```{r}
#Define model and priors
# Define the formula

AltercentricDiagnosis_f0 <- bf(
  AltercentricIntrusion ~ 1 + Diagnosis
)

AltercentricDiagnosis_f <- bf(
  AltercentricIntrusion ~ 0 + Diagnosis
)

# Design the priors

get_prior(AltercentricDiagnosis_f0, family = gaussian, d) #Intercept prior = expectation of average level of altercentric intrusion in controls
                                                          #Beta prior = expectation of difference between schizo and control in AI 
                                                          #Sigma prior = expectation of the average error
get_prior(AltercentricDiagnosis_f, family = gaussian, d)

#We inspect the data to define a prior
summary(d$AltercentricIntrusion)
sd(d$AltercentricIntrusion)
#we decide on a mean of 4 and a sd of 1 based on the summary
priorDiagnosis <- c(
  prior(normal(4, 1), class = b),
  prior(normal(1, 2), class = sigma)
) 

#Prior checks
AltercentricDiagnosis_PriorCheck_m <- brm(
  formula = AltercentricDiagnosis_f,
  data = d,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = "only" #does not take data into account
)

#Prior predictive checks
pp_check(AltercentricDiagnosis_PriorCheck_m, nsamples = 100)

## Fitting the model
AltercentricDiagnosis_m <- brm(
  formula = AltercentricDiagnosis_f,
  data = d,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = T #takes data into account
)

#Posterior predictive checks
pp_check(AltercentricDiagnosis_m, nsamples = 100)

## Check the model for warnings
AltercentricDiagnosis_m

#Plots
# Hypothesis testing + updating check
plot(hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > Intercept"))

hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > Intercept")
#4000 samples define our posterior
#Evidence ratio = if you take the amount

conditional_effects(AltercentricDiagnosis_m)

plot(conditional_effects(AltercentricDiagnosis_m), points=T)

```


Q1.2) Is altercentric intrusion related to specific symptoms *in the patients*? Identify which of the symptoms could be relevant. Should you include more than one symptom? Build models, priors, predictive checks. Assess the evidence and report models and results, including plots. Discuss whether the results make sense.

All the univariate models
```{r}
#Isolate the patients
d2 <- filter(d,Diagnosis == "Schizophrenia")

##Model the three univariate models##
plot(AltercentricIntrusion~VoiceHearing, d2)

## Voice hearing ##

# Define the formula
Altercentric_VH_f <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing
)
# Design the priors
get_prior(Altercentric_VH_f, family = gaussian, d2)

#Choosing the priors
summary(d2$AltercentricIntrusion)
sd(d2$AltercentricIntrusion)

priorDiagnosis <- c(
  prior(normal(3.5, 1), class = Intercept),
  prior(normal(0.5,1), class = b), # for beta
  prior(normal(1, 2), class = sigma)
) 

#Prior checks
Altercentric_VH_PriorCheck_m <- brm(
  formula = Altercentric_VH_f,
  data = d2,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = "only" #does not take data into account
)

#Prior predictive checks
pp_check(Altercentric_VH_PriorCheck_m, nsamples = 100)

## Fitting the model
Altercentric_VH_m <- brm(
  formula = Altercentric_VH_f,
  data = d2,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = T #takes data into account
)

#Posterior predictive checks
pp_check(Altercentric_VH_m, nsamples = 100)
Altercentric_VH_m

#there is no effect of voice hearing

## Mind reading ##

# Define the formula
Altercentric_MR_f <- bf(
  AltercentricIntrusion ~ 1 + MindReading
)
# Design the priors
get_prior(Altercentric_MR_f, family = gaussian, d2)

#Choosing the priors
summary(d2$AltercentricIntrusion)
sd(d2$AltercentricIntrusion)

priorDiagnosis <- c(
  prior(normal(3.5, 1), class = Intercept),
  prior(normal(0.5,1), class = b), # for beta
  prior(normal(1, 2), class = sigma)
) 

#Prior checks
Altercentric_MR_PriorCheck_m <- brm(
  formula = Altercentric_MR_f,
  data = d2,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = "only" #does not take data into account
)

#Prior predictive checks
pp_check(Altercentric_MR_PriorCheck_m, nsamples = 100)

## Fitting the model
Altercentric_MR_m <- brm(
  formula = Altercentric_MR_f,
  data = d2,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = T #takes data into account
)

#Posterior predictive checks
pp_check(Altercentric_MR_m, nsamples = 100)
Altercentric_MR_m
#there is no effect of mind reading

## Apathy ##

# Define the formula
Altercentric_Apathy_f <- bf(
  AltercentricIntrusion ~ 1 + Apathy
)
# Design the priors
get_prior(Altercentric_Apathy_f, family = gaussian, d2)

#Choosing the priors
summary(d2$AltercentricIntrusion)
sd(d2$AltercentricIntrusion)

priorDiagnosis <- c(
  prior(normal(4.5, 1), class = Intercept),
  prior(normal(-0.5,1), class = b), # for beta
  prior(normal(1, 2), class = sigma)
) 

#Prior checks
Altercentric_Apathy_PriorCheck_m <- brm(
  formula = Altercentric_Apathy_f,
  data = d2,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = "only" #does not take data into account
)

#Prior predictive checks
pp_check(Altercentric_Apathy_PriorCheck_m, nsamples = 100)

## Fitting the model
Altercentric_Apathy_m <- brm(
  formula = Altercentric_Apathy_f,
  data = d2,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = T #takes data into account
)

#Posterior predictive checks
pp_check(Altercentric_Apathy_m, nsamples = 100)
Altercentric_Apathy_m
```

Voice hearing and mind reading model
```{r}
##Model the multivariate regression##
#mean-centering
d2 <- d2 %>% 
  mutate(
    VoiceHearing_s = VoiceHearing - mean(VoiceHearing),
    MindReading_s = MindReading - mean(MindReading),
    Apathy_s = Apathy - mean(Apathy)
  )

round(mean(d2$VoiceHearing_s),3)

#Multivariate VH and MR 
VH_MR_f <- bf(
  AltercentricIntrusion ~ 1 + MindReading_s + VoiceHearing_s
)
get_prior(VH_MR_f, d2)
mean(d2$AltercentricIntrusion)

#Define priors
priorList <- c(
  prior(normal(4, 1), class = Intercept),
  prior(normal(0.1,1), class = b), # for beta
  prior(normal(1, 2), class = sigma)
) 

#Prior checks
VH_MR_PriorCheck <- brm(
  formula = VH_MR_f,
  data = d2,
  family = gaussian,
  prior = priorList,
  sample_prior = "only" #does not take data into account
)

#Prior predictive checks
pp_check(VH_MR_PriorCheck, nsamples = 100)

## Fitting the model
VH_MR_m <- brm(
  formula = VH_MR_f,
  data = d2,
  family = gaussian,
  prior = priorList,
  sample_prior = T #takes data into account
)

#Posterior predictive checks
pp_check(VH_MR_m, nsamples = 100)
VH_MR_m
#plots...

```

Apathy and Voice hearing model
```{r}
#Multivariate VH and Apathy 
VH_A_f <- bf(
  AltercentricIntrusion ~ 1 + Apathy_s + VoiceHearing_s
)
get_prior(VH_A_f, d2)
mean(d2$AltercentricIntrusion)

#Define priors
priorList <- c(
  prior(normal(4, 1), class = Intercept),
  prior(normal(-0.5,1), class = b, coef = Apathy_s), # for beta
  prior(normal(0.5,1), class = b, coef = VoiceHearing_s), # for beta
  prior(normal(1, 2), class = sigma)
) 

#Prior checks
VH_A_PriorCheck <- brm(
  formula = VH_A_f,
  data = d2,
  family = gaussian,
  prior = priorList,
  sample_prior = "only" #does not take data into account
)

#Prior predictive checks
pp_check(VH_A_PriorCheck, nsamples = 100)

## Fitting the model
VH_A_m <- brm(
  formula = VH_A_f,
  data = d2,
  family = gaussian,
  prior = priorList,
  sample_prior = T #takes data into account
)

#Posterior predictive checks
pp_check(VH_A_m, nsamples = 100)
print(VH_A_m, digits = 3)
```

All predictors
```{r}
# Define the formula
VoiceMindApathy <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing_s + MindReading_s + Apathy_s)
# Design the priors
get_prior(VoiceMindApathy, family = gaussian, d2)
prior_VoiceMindApathy <- c(
  prior(normal(4, 1), class = Intercept), # It is a good guess that when voice hearing and mind reading er mean, 
  prior(normal(0.1, 1), class = b, coef = VoiceHearing_s),# så er AI også
  prior(normal(0.1, 1), class = b, coef = MindReading_s),
  prior(normal(-0.5, 1), class = b, coef = Apathy_s),
  prior(normal(1, 2), class = sigma) # average error we expect
) 
# Test the priors
VoiceMindApathy_m <- brm(
  formula = VoiceMindApathy,
  data = d2,
  family = gaussian,
  prior = prior_VoiceMindApathy,
  sample_prior = "only" # meaning we don't take the data into account
)
# check how the prediction given only the priors and not the data will look like 
pp_check(VoiceMindApathy_m, nsamples = 100)
## Fitting the model
VoiceMindApathy_m2 <- brm(
  formula = VoiceMindApathy,
  data = d2,
  family = gaussian,
  prior = prior_VoiceMindApathy,
  sample_prior = T # only change, now we take data into account
)
# Posterior predictive check
pp_check(VoiceMindApathy_m2, nsamples = 100)
## Check the model for warnings
VoiceMindApathy_m2

```


```{r}
#Plots
# Hypothesis testing + updating check
plot(hypothesis(VoiceMindApathy_m2,
           "Apathy_s > Intercept",
          ))

hypothesis(VoiceMindApathy_m2,
           "Apathy_s > Intercept")

?hypothesis
#4000 samples define our posterior
#Evidence ratio = if you take the amount
?conditional_effects()

conditional_effects(Altercentric_Apathy_m)

plot(conditional_effects(VoiceMindApathy_m2), points=T)
```


## Second part

Q2.1) However, we know that the diagnosis is based on symptom assessment: if the overall sum of symptoms is severe enough, the participant gets a diagnosis. In other words, by selecting the patients, and including the symptoms in the model we might have inadvertently introduced an issue in our inference. Do try to draw a causal graph (Directed Acyclical Graph) of the variables and compare it with the types of causal graphs presented in the slides. Discuss which biases you might have introduced.
```{r}
#Causal graph
```


Q2.2.) Redesign your analysis following the graph and report how the results change
```{r}
#What should we do to avoid the issues that we infer from the graph
```


## Third part

These issues are very difficult to think through, and not knowing the causal mechanisms generating the data in advance makes our inferences even more unreliable. To explore these issues, I recommend using simulations. In other words, defining a "true" model, generating data from it and assessing what different analyses would lead you to infer (and therefore which biases they might introduce). You can find the code I used to simulate your data below.

Q3.1) Look through the code and identify whether the results you have match the underlying truth. Discuss what you have learned.

Q3.2) OPTIONAL: is this a general pattern? Try varying the parameters (e.g. correlation values) and assess whether the new dataset(s) leads to the same biases in your analysis.



```{r}
pacman::p_load(MASS, tidyverse, psych)

seed <- 1981 # Defining a seed so the results are always the same
n <- 300 # Defining the amount of participants

SymptomCorr <- .2 # Defining the correlation of symptoms (as they tend to co-occur)
EffectCorrRel <- .2 # Defining the correlation between relevant symptoms and effect (Some symptoms are positively correlated with the effect)
EffectCorrIrrel <- 0 # Defining the correlation between irrelevant symptoms and effect (none)

# Creating the variance-covariance matrix for the variables we want to generate (3 symptoms, 1 effect)
Sigma <- matrix(data=c(1,SymptomCorr,SymptomCorr,EffectCorrRel,
                       SymptomCorr,1,SymptomCorr,EffectCorrRel,
                       SymptomCorr,SymptomCorr,1,EffectCorrIrrel,
                       EffectCorrRel,EffectCorrRel,EffectCorrIrrel,1),
                       nrow=4,ncol=4)

## Generate data from a multivariate (mvr) normal (n) distribution
d <- mvrnorm(n = n, # number of participant
        mu = c(1.2, 1.2, 1.2, 4), # mean of each variable
        Sigma) # variance co-variance matrix

# Giving meaningful names to variables and add ID
d <- data.frame(
  VoiceHearing = d[,1], 
  MindReading =  d[,2],
  Apathy =  d[,3], 
  AltercentricIntrusion = d[,4],
  ID = seq(nrow(d)))

# Assessing whether the participant has schizophrenia (high enough sum of symptoms)
# Here we choose participants scoring above 75% percentile (the most severe ones)
d$Diagnosis <- 0
d$Diagnosis[(d$VoiceHearing + d$MindReading + d$Apathy) > 
              quantile(d$VoiceHearing + d$MindReading + d$Apathy, .75)] <-1

## Plotting the relation between variables in schizophrenia
d1 <- d %>% subset(Diagnosis==1) %>% dplyr::select(-Diagnosis, -ID)
pairs.panels(d1)

## Plotting the relation between variables all participants
pairs.panels(dplyr::select(d,-Diagnosis, -ID))

write_csv(d, "data/Ass3.csv")
```

